Weighted Model Counting Without Parameter Variables

### 2021-06-19 doc/SAT_long_talk/talk.tex
Update some details

\begin{frame}
  \tikz[remember picture,overlay]{
    \node at ([yshift=35pt,xshift=30pt]current page.south)
    {\includegraphics[height=40pt]{../poster/logo_inf.png}};
    \node at ([yshift=35pt,xshift=75pt]current page.south)
    {\includegraphics[height=40pt]{../poster/logo_ecr.png}};
    \node at ([yshift=30pt,xshift=140pt]current page.south)
    {\includegraphics[height=20pt]{../poster/logo_ukri.png}};
  }
  \titlepage
\end{frame}

\begin{frame}[fragile]{The Computational Problem of Probabilistic Inference}
  \vspace{-0.75cm}
  \begin{columns}[t]
    \begin{column}{0.65\textwidth}
      \centering
      \begin{block}{ProbLog}
        \vspace{-0.2cm}
        \begin{lstlisting}[basicstyle=\tiny]
0.001 :: burglary.
0.002 :: earthquake.
0.95  :: alarm     :- burglary, earthquake.
0.94  :: alarm     :- burglary, \+ earthquake.
0.29  :: alarm     :- \+ burglary, earthquake.
0.001 :: alarm     :- \+ burglary, \+ earthquake.
0.9   :: johnCalls :- alarm.
0.05  :: johnCalls :- \+ alarm.
0.7   :: maryCalls :- alarm.
0.01  :: maryCalls :- \+ alarm.
        \end{lstlisting}
        \vspace{-0.2cm}
      \end{block}
      \vspace{-0.25cm}
      \begin{block}{BLOG}
        \vspace{-0.2cm}
        \begin{lstlisting}[escapeinside={(*}{*)},basicstyle=\tiny]
random Boolean Burglary (*$\sim$*) BooleanDistrib(0.001);
random Boolean Earthquake (*$\sim$*) BooleanDistrib(0.002);
random Boolean Alarm (*$\sim$*)
  if Burglary then
    if Earthquake then BooleanDistrib(0.95)
    else BooleanDistrib(0.94)
  else
    if Earthquake then BooleanDistrib(0.29)
    else BooleanDistrib(0.001);
random Boolean JohnCalls (*$\sim$*)
  if Alarm then BooleanDistrib(0.9)
  else BooleanDistrib(0.05);
random Boolean MaryCalls (*$\sim$*)
  if Alarm then BooleanDistrib(0.7)
  else BooleanDistrib(0.01);
        \end{lstlisting}
        \vspace{-0.2cm}
      \end{block}
    \end{column}
    \begin{column}{0.35\textwidth}
      \begin{block}{Bayesian Network}
        \centering
        \begin{tikzpicture}[node distance=2cm,scale=0.5,every node/.style={scale=0.5}]
          \node[draw,ellipse] (alarm) {Alarm};
          \node[draw,ellipse,above left of=alarm] (burglary) {Burglary};
          \node[draw,ellipse,above right of=alarm] (earthquake) {Earthquake};
          \node[draw,ellipse,below left of=alarm] (johnCalls) {JohnCalls};
          \node[draw,ellipse,below right of=alarm] (maryCalls) {MaryCalls};
          \draw[-Latex] (burglary) -- (alarm);
          \draw[-Latex] (earthquake) -- (alarm);
          \draw[-Latex] (alarm) -- (johnCalls);
          \draw[-Latex] (alarm) -- (maryCalls);
        \end{tikzpicture}
      \end{block}
      \vspace{1cm}
      \begin{block}{Markov Random Field}
        \centering
        \begin{tikzpicture}[node distance=2cm,scale=0.5,every node/.style={scale=0.5}]
          \node[draw,ellipse] (alarm) {Alarm};
          \node[draw,ellipse,above left of=alarm] (burglary) {Burglary};
          \node[draw,ellipse,above right of=alarm] (earthquake) {Earthquake};
          \node[draw,ellipse,below left of=alarm] (johnCalls) {JohnCalls};
          \node[draw,ellipse,below right of=alarm] (maryCalls) {MaryCalls};
          \draw (burglary) -- (earthquake);
          \draw (burglary) -- (alarm);
          \draw (earthquake) -- (alarm);
          \draw (alarm) -- (johnCalls);
          \draw (alarm) -- (maryCalls);
        \end{tikzpicture}
      \end{block}
    \end{column}
  \end{columns}
  \onslide<2>{
    \begin{tikzpicture}[remember picture,overlay]
      \node[draw,star,fill=red!10] (wmc) at (current page.center) {WMC};
      \coordinate[xshift=-0.25\linewidth,yshift=-0.25\textheight] (p1) at (current page.center);
      \coordinate[xshift=0.25\linewidth,yshift=-0.25\textheight] (p2) at (current page.center);
      \coordinate[xshift=-0.25\linewidth,yshift=0.25\textheight] (p3) at (current page.center);
      \coordinate[xshift=0.25\linewidth,yshift=0.25\textheight] (p4) at (current page.center);
      \draw[-latex,line width=2pt,color=red!50] (p1) -- (wmc);
      \draw[-latex,line width=2pt,color=red!50] (p2) -- (wmc);
      \draw[-latex,line width=2pt,color=red!50] (p3) -- (wmc);
      \draw[-latex,line width=2pt,color=red!50] (p4) -- (wmc);
    \end{tikzpicture}
  }
\end{frame}

@@

\begin{frame}{WMC, Formally}
  \begin{definition}
    A \alert{WMC instance} is a tuple \structure{$(\phi, X_I, X_P, w)$}, where
    \begin{itemize}
    \item \structure{$X_I$} is the set of \alert{indicator variables},
    \item \structure{$X_P$} is the set of \alert{parameter variables} (with
      \structure{$X_I \cap X_P = \emptyset$}),
    \item \structure{$\phi$} is a propositional formula in CNF over
      \structure{$X_I \cup X_P$},
    \item \structure{$w\colon X_I \cup X_P \cup \{\neg x \mid x \in X_I \cup
        X_P\} \to \mathbb{R}$} is a \alert{weight function}
      \begin{itemize}
      \item such that \structure{$w(x) = w(\neg x) = 1$} for all \structure{$x
          \in X_I$}.
      \end{itemize}
    \end{itemize}
  \end{definition}
  \begin{definition}
    Let \structure{$\phi$} be a formula over a set of variables \structure{$X$}.
    Then \structure{$Y \subseteq X$} is a \alert{minimum-cardinality model} of
    \structure{$\phi$} if
    \begin{itemize}
    \item \structure{$Y \models \phi$},
    \item and \structure{$|Y| \le |Z|$} for all \structure{$Z \models \phi$}.
    \end{itemize}
  \end{definition}
\end{frame}

\begin{frame}{WMC and Minimum-Cardinality WMC}
  The goal of \alert{WMC} is to compute
  \[
    \sum_{Y \models \phi} \prod_{Y \models l} w(l)
  \]
  whereas the goal of \alert{minimum-cardinality WMC} is to compute
  \[
    \sum_{Y \models \phi\text{, }\alert{|Y| = k}} \prod_{Y \models l} w(l),
  \]
  where
  \[
    k = \min_{Y \models \phi} |Y|.
  \]
\end{frame}

\begin{frame}{A More Expressive Alternative}
  For any propositional formula \structure{$\phi$} over a set of variables
  \structure{$X$} and \structure{$p, q \in \mathbb{R}$}, let
  \structure{$[\phi]^p_q\colon 2^X \to \mathbb{R}$} be the pseudo-Boolean
  function defined as
  \[
    [\phi]^p_q(Y) \coloneqq
    \begin{cases}
      p & \text{if } Y \models \phi \\
      q & \text{otherwise}
    \end{cases}
  \]
  for any \structure{$Y \subseteq X$}.

  \begin{definition}[Pseudo-Boolean Projection (PBP)]
    A \alert{PBP instance} is a tuple \structure{$(F, X, \omega)$}, where
    \structure{$X$} is the set of variables, \structure{$F$} is a set of
    two-valued pseudo-Boolean functions \structure{$2^X \to \mathbb{R}$}, and
    \structure{$\omega \in \mathbb{R}$} is the scaling factor.
  \end{definition}
\end{frame}

\begin{frame}{From WMC to PBP}
  \begin{example}
    \begin{itemize}
    \item Indicator variable: \structure{$x$}
    \item Parameter variables: \structure{$p$}, \structure{$q$}
    \item Weights: \structure{$w(p) = 0.2$}, \structure{$w(q) = 0.8$}, and \structure{$w(\neg p) = w(\neg q) = 1$}
    \end{itemize}
    \begin{center}
      \begin{tabular}{llll}
        \toprule
        WMC Clause & \onslide<2->{In CNF} & \onslide<3->{Pseudo-Boolean Function} & \\
        \midrule
        $\neg x \Rightarrow p$ & \onslide<2->{$x \lor p$} & \onslide<3->{$[\neg x]_1^{0.2}$} & \\
        $p \Rightarrow \neg x$ & \onslide<2->{$\neg x \lor \neg p$} & & \onslide<4->{$[x]^{0.8}_{0.2}$} \\
        $x \Rightarrow q$ & \onslide<2->{$\neg x \lor q$} & \onslide<3->{$[x]_1^{0.8}$} & \\
        $q \Rightarrow x$ & \onslide<2->{$x \lor \neg q$} & & \\
        $\neg x$ & \onslide<2->{$\neg x$} & \onslide<3->{$[\neg x]_0^1$} & \onslide<4->{$[\neg x]_0^1$} \\
        \bottomrule
      \end{tabular}
    \end{center}
  \end{example}
\end{frame}

@@

\begin{frame}{Correctness Conditions for WMC (1/2)}
  For each parameter variable \structure{$p \in X_P$},
  \begin{itemize}
  \item \structure{$w(\neg p) = 1$},
  \item and the set of clauses that mention \structure{$p$} or \structure{$\neg
      p$} is
    \[
      \left\{p \lor \bigvee_{i=1}^n \neg l_i \right\} \cup \{ l_i \lor \neg p
      \mid i = 1, \dots, n \}
    \]
    for some non-empty family of \alert{indicator} literals
    \structure{$(l_i)_{i=1}^n$}.
  \end{itemize}
\end{frame}

\begin{frame}{Correctness Conditions for WMC (2/2)}
  For each parameter variable \structure{$p \in X_P$},
  \begin{itemize}
  \item \structure{$w(p) + w(\neg p) = 1$},
  \item each clause has at most one parameter variable,
  \item there is no clause \structure{$c \in \phi$} such that \structure{$\neg p
      \in c$},
  \item if \structure{$\{ p \} \in \phi$}, then this is the only clause that
    mentions \structure{$p$},
  \item and for any \structure{$c, d \in \phi$} such that \structure{$c \ne d$},
    \structure{$p \in c$} and \structure{$p \in d$},
    \[
      \bigwedge_{l \in c \setminus \{ p \}} \neg l \land \bigwedge_{l \in d
        \setminus \{ p \}} \neg l
    \]
    is false.
  \end{itemize}
\end{frame}

\begin{frame}{Additional Conditions for Minimum-Cardinality WMC}
  \begin{itemize}
  \item All models of \structure{$\{ c \in \phi \mid c \cap X_P = \emptyset \}$}
    have the same number of positive indicator literals,
  \item and
    \[
      \min_{Z \subseteq X_P} |Z| \quad \text{s.t. } \quad Y \cup Z \models \phi
    \]
    is the same for all \structure{$Y \models \{ c \in \phi \mid c \cap X_P =
      \emptyset \}$}.
  \end{itemize}
\end{frame}

@@

\begin{frame}{WMC/PBP Encodings for Bayesian Networks}
  \centering
  \includegraphics[width=\textwidth]{cumulative5}
\end{frame}

@@

\section{Summary}

\begin{frame}{Summary}
  \begin{itemize}
  \item TODO
  \end{itemize}
\end{frame}

>>>>

\begin{frame}
  \tikz[remember picture,overlay]{
    \node at ([yshift=20pt,xshift=30pt]current page.south)
    {\includegraphics[height=40pt]{../poster/logo_inf.png}};
    \node at ([yshift=20pt,xshift=75pt]current page.south)
    {\includegraphics[height=40pt]{../poster/logo_ecr.png}};
    \node at ([yshift=15pt,xshift=140pt]current page.south)
    {\includegraphics[height=20pt]{../poster/logo_ukri.png}};
  }
  \titlepage
\end{frame}

\begin{frame}[fragile]{The Computational Problem of Probabilistic Inference}
  \vspace{-0.75cm}
  \begin{columns}[t]
    \begin{column}{0.65\textwidth}
      \centering
      \begin{block}{ProbLog}
        \vspace{-0.2cm}
        \begin{lstlisting}[basicstyle=\tiny]
0.001 :: burglary.
0.002 :: earthquake.
0.95  :: alarm     :- burglary, earthquake.
0.94  :: alarm     :- burglary, \+ earthquake.
0.29  :: alarm     :- \+ burglary, earthquake.
0.001 :: alarm     :- \+ burglary, \+ earthquake.
0.9   :: johnCalls :- alarm.
0.05  :: johnCalls :- \+ alarm.
0.7   :: maryCalls :- alarm.
0.01  :: maryCalls :- \+ alarm.
        \end{lstlisting}
        \vspace{-0.2cm}
      \end{block}
      \begin{block}{BLOG}
        \vspace{-0.2cm}
        \begin{lstlisting}[escapeinside={(*}{*)},basicstyle=\tiny]
random Boolean Burglary (*$\sim$*) BooleanDistrib(0.001);
random Boolean Earthquake (*$\sim$*) BooleanDistrib(0.002);
random Boolean Alarm (*$\sim$*)
  if Burglary then
    if Earthquake then BooleanDistrib(0.95)
    else BooleanDistrib(0.94)
  else
    if Earthquake then BooleanDistrib(0.29)
    else BooleanDistrib(0.001);
random Boolean JohnCalls (*$\sim$*)
  if Alarm then BooleanDistrib(0.9)
  else BooleanDistrib(0.05);
random Boolean MaryCalls (*$\sim$*)
  if Alarm then BooleanDistrib(0.7)
  else BooleanDistrib(0.01);
        \end{lstlisting}
        \vspace{-0.2cm}
      \end{block}
    \end{column}
    \begin{column}{0.35\textwidth}
      \begin{block}{Bayesian Network}
        \centering
        \begin{tikzpicture}[node distance=2cm,scale=0.5,every node/.style={scale=0.5}]
          \node[draw,ellipse] (alarm) {Alarm};
          \node[draw,ellipse,above left of=alarm] (burglary) {Burglary};
          \node[draw,ellipse,above right of=alarm] (earthquake) {Earthquake};
          \node[draw,ellipse,below left of=alarm] (johnCalls) {JohnCalls};
          \node[draw,ellipse,below right of=alarm] (maryCalls) {MaryCalls};
          \draw[-Latex] (burglary) -- (alarm);
          \draw[-Latex] (earthquake) -- (alarm);
          \draw[-Latex] (alarm) -- (johnCalls);
          \draw[-Latex] (alarm) -- (maryCalls);
        \end{tikzpicture}
      \end{block}
      \vspace{1cm}
      \begin{block}{Markov Random Field}
        \centering
        \begin{tikzpicture}[node distance=2cm,scale=0.5,every node/.style={scale=0.5}]
          \node[draw,ellipse] (alarm) {Alarm};
          \node[draw,ellipse,above left of=alarm] (burglary) {Burglary};
          \node[draw,ellipse,above right of=alarm] (earthquake) {Earthquake};
          \node[draw,ellipse,below left of=alarm] (johnCalls) {JohnCalls};
          \node[draw,ellipse,below right of=alarm] (maryCalls) {MaryCalls};
          \draw (burglary) -- (earthquake);
          \draw (burglary) -- (alarm);
          \draw (earthquake) -- (alarm);
          \draw (alarm) -- (johnCalls);
          \draw (alarm) -- (maryCalls);
        \end{tikzpicture}
      \end{block}
    \end{column}
  \end{columns}
  \onslide<2>{
    \begin{tikzpicture}[remember picture,overlay]
      \node[draw,star,fill=red!10] (wmc) at (current page.center) {WMC};
      \coordinate[xshift=-0.25\linewidth,yshift=-0.25\textheight] (p1) at (current page.center);
      \coordinate[xshift=0.25\linewidth,yshift=-0.25\textheight] (p2) at (current page.center);
      \coordinate[xshift=-0.25\linewidth,yshift=0.25\textheight] (p3) at (current page.center);
      \coordinate[xshift=0.25\linewidth,yshift=0.25\textheight] (p4) at (current page.center);
      \draw[-latex,line width=2pt,color=red!50] (p1) -- (wmc);
      \draw[-latex,line width=2pt,color=red!50] (p2) -- (wmc);
      \draw[-latex,line width=2pt,color=red!50] (p3) -- (wmc);
      \draw[-latex,line width=2pt,color=red!50] (p4) -- (wmc);
    \end{tikzpicture}
  }
\end{frame}

@@

\begin{frame}{WMC, Formally}
  \begin{definition}
    A \alert{WMC instance} is a tuple \structure{$(\phi, X_I, X_P, w)$}, where
    \begin{itemize}
    \item \structure{$X_I$} is the set of \alert{indicator variables},
    \item \structure{$X_P$} is the set of \alert{parameter variables} (with
      \structure{$X_I \cap X_P = \emptyset$}),
    \item \structure{$\phi$} is a propositional formula in CNF over
      \structure{$X_I \cup X_P$},
    \item \structure{$w\colon X_I \cup X_P \cup \{\neg x \mid x \in X_I \cup
        X_P\} \to \mathbb{R}$} is a \alert{weight function}
      \begin{itemize}
      \item such that \structure{$w(x) = w(\neg x) = 1$} for all \structure{$x
          \in X_I$}.
      \end{itemize}
    \end{itemize}
  \end{definition}
  % \begin{definition}
  %   Let \structure{$\phi$} be a formula over a set of variables \structure{$X$}.
  %   Then \structure{$Y \subseteq X$} is a \alert{minimum-cardinality model} of
  %   \structure{$\phi$} if
  %   \begin{itemize}
  %   \item \structure{$Y \models \phi$},
  %   \item and \structure{$|Y| \le |Z|$} for all \structure{$Z \models \phi$}.
  %   \end{itemize}
  % \end{definition}
\end{frame}

% \begin{frame}{WMC and Minimum-Cardinality WMC}
%   The goal of \alert{WMC} is to compute
%   \[
%     \sum_{Y \models \phi} \prod_{Y \models l} w(l)
%   \]
%   whereas the goal of \alert{minimum-cardinality WMC} is to compute
%   \[
%     \sum_{Y \models \phi\text{, }\alert{|Y| = k}} \prod_{Y \models l} w(l),
%   \]
%   where
%   \[
%     k = \min_{Y \models \phi} |Y|.
%   \]
% \end{frame}

\begin{frame}{A More Expressive Alternative}
  \begin{definition}[Pseudo-Boolean Projection (PBP)]
    A \alert{PBP instance} is a tuple \structure{$(F, X, \omega)$}, where
    \structure{$X$} is the set of variables, \structure{$F$} is a set of
    two-valued pseudo-Boolean functions \structure{$2^X \to \mathbb{R}$}, and
    \structure{$\omega \in \mathbb{R}$} is the scaling factor.
  \end{definition}

  For any propositional formula \structure{$\phi$} over a set of variables
  \structure{$X$} and \structure{$p, q \in \mathbb{R}$}, let
  \structure{$[\phi]^p_q\colon 2^X \to \mathbb{R}$} be the pseudo-Boolean
  function defined as
  \[
    [\phi]^p_q(Y) \coloneqq
    \begin{cases}
      p & \text{if } Y \models \phi \\
      q & \text{otherwise}
    \end{cases}
  \]
  for any \structure{$Y \subseteq X$}.
\end{frame}

\begin{frame}{From WMC to PBP}
  \begin{example}
    \begin{itemize}
    \item Indicator variable: \structure{$x$}
    \item Parameter variables: \structure{$p$}, \structure{$q$}
    \item Weights: \structure{$w(p) = 0.2$}, \structure{$w(q) = 0.8$}, and \structure{$w(\neg p) = w(\neg q) = 1$}
    \end{itemize}
    \begin{center}
      \begin{tabular}{llll}
        \toprule
        WMC Clause & \onslide<2->{In CNF} & \onslide<3->{Pseudo-Boolean Function} & \\
        \midrule
        $\neg x \Rightarrow p$ & \onslide<2->{$x \lor p$} & \onslide<3->{$[\neg x]_1^{0.2}$}\tikz \coordinate (one); & \\
        $p \Rightarrow \neg x$ & \onslide<2->{$\neg x \lor \neg p$} & & \tikz \coordinate (two); \onslide<4->{$[x]^{0.8}_{0.2}$} \\
        $x \Rightarrow q$ & \onslide<2->{$\neg x \lor q$} & \onslide<3->{$[x]_1^{0.8}$} \tikz \coordinate (three); & \\
        $q \Rightarrow x$ & \onslide<2->{$x \lor \neg q$} & & \\
        $\neg x$ & \onslide<2->{$\neg x$} & \onslide<3->{$[\neg x]_0^1$} & \onslide<4->{$[\neg x]_0^1$} \\
        \bottomrule
      \end{tabular}
    \end{center}
  \end{example}
  \onslide<4->{
    \begin{tikzpicture}[remember picture,overlay]
      \draw[-Latex,blue] (one) -- (two);
      \draw[-Latex,blue] (three) -- (two);
    \end{tikzpicture}
  }
\end{frame}

@@

\begin{frame}{Correctness Conditions (1/2)}
  For each parameter variable \structure{$p \in X_P$},
  \begin{itemize}
  \item \structure{$w(\neg p) = 1$},
  \item and the set of clauses that mention \structure{$p$} or \structure{$\neg
      p$} is
    \[
      \left\{p \lor \bigvee_{i=1}^n \neg l_i \right\} \cup \{ l_i \lor \neg p
      \mid i = 1, \dots, n \}
    \]
    for some non-empty family of \alert{indicator} literals
    \structure{$(l_i)_{i=1}^n$}.
  \end{itemize}
  In other words, \structure{$p$} is defined to be equivalent to
  \structure{$\bigwedge_{i=1}^n l_i$}.
\end{frame}

\begin{frame}{Correctness Conditions (2/2)}
  For each parameter variable \structure{$p \in X_P$},
  \begin{itemize}
  \item \structure{$w(p) + w(\neg p) = 1$},
  \item each clause has at most one parameter variable,
  \item there are no negative parameter literals,
  \item if \structure{$\{ p \} \in \phi$}, then this is the only clause that
    mentions \structure{$p$},
  \item and for any two clauses of the form \structure{$\chi \Rightarrow p$} and
    \structure{$\psi \Rightarrow p$}, \structure{$\chi \land \psi \equiv \bot$}.
  \end{itemize}
\end{frame}

% \begin{frame}{Additional Conditions for Minimum-Cardinality WMC}
%   \begin{itemize}
%   \item All models of \structure{$\{ c \in \phi \mid c \cap X_P = \emptyset \}$}
%     have the same number of positive indicator literals,
%   \item and
%     \[
%       \min_{Z \subseteq X_P} |Z| \quad \text{s.t. } \quad Y \cup Z \models \phi
%     \]
%     is the same for all \structure{$Y \models \{ c \in \phi \mid c \cap X_P =
%       \emptyset \}$}.
%   \end{itemize}
% \end{frame}

@@

@@

\section{Summary and Future Work}

\begin{frame}{Summary and Future Work}
  \begin{itemize}
  \item PBP is a more expressive alternative to WMC that works with
    state-of-the-art WMC algorithms based on pseudo-Boolean function
    manipulation.
  \item Many WMC encodings can be efficiently transformed into PBP while
    removing unnecessary variables and clauses.
  \item The identified conditions for this transformation to work help explain
    how WMC encodings for Bayesian networks operate.
  \item Performance improvements depend on the encoding.
    \begin{itemize}
    \item The very first encoding was virtually unaffected,
    \item whereas the state-of-the-art encoding was significantly improved.
    \end{itemize}
  \item Can the identified conditions be generalised further?
  \item Can the transformation be applied to WMC encodings for other application
    domains?
  \end{itemize}
\end{frame}
